---
title: mlr3spatial
author: Marc Becker
date: '2021-12-01'
slug: mlr3spatial
categories:
  - R
tags:
  - mlr3spatial
  - CRAN
---

We are happy to announce that [mlr3spatial](https://mlr3spatial.mlr-org.com/) has been released on [CRAN](https://cran.r-project.org/web/packages/mlr3spatial/index.html) in November 2021. 
[mlr3spatial](https://mlr3spatial.mlr-org.com/) simplifies the handling of spatial objects in the [mlr3](https://github.com/mlr-org) ecosystem.
Before [mlr3spatial](https://mlr3spatial.mlr-org.com/), the user had to tabular data from spatial objects to train a model or predict a map.

Now, mlr3 [tasks](https://mlr3.mlr-org.com/reference/Task.html) can directly read from spatial objects via specialized [data backends](https://mlr3.mlr-org.com/reference/DataBackend.html).
Such tasks can be used to train a model or to perform resampling just like any other mlr3 task.
We support spatial raster objects created by the [terra](https://CRAN.R-project.org/package=terra), [raster](https://CRAN.R-project.org/package=raster) and [stars](https://CRAN.R-project.org/package=stars) packages.
Additionally, vector data is handled with the package [sf](https://CRAN.R-project.org/package=sf).

The `predict_raster()` function makes a spatial raster with predictions from an mlr3 learner.
We only have to pass a task with a spatial data backend which provides the data and spatial reference.
To avoid memory issues with large raster files, the prediction is done in chunks.
One after the other, the raster chunks are loaded into memory and the prediction is written to disk.
Finally, the complete raster is available on disk. 
The learner can also make use of future-based parallelization to accelerate the predictions.
The vignette on [“Benchmarking parallel predictions”](https://mlr3spatial.mlr-org.com/articles/benchmark.html) showcases the parallelization capabilities of [mlr3spatial](https://mlr3spatial.mlr-org.com/).

## Use Case - Landsat7 data as {stars} object

### Data Preparation

```{r, include=FALSE}
library("stars")
```

```{r}
library("mlr3")
library("mlr3spatial")
```

First, the TIFF files is read via `stars::read_stars()` and put into a `DataBackendRaster`.
The DataBackend is then used to create a regression task with the response being `layer.1`.

```{r, error=TRUE}
tif = system.file("tif/L7_ETMs.tif", package = "stars")
stack = stars::read_stars(tif)

backend = as_data_backend(stack)
task = as_task_regr(backend, target = "layer.1")

print(task)
```

For large raster files with millions of values it helps to predict in parallel.
To enable this, set `learner$parallel_predict = TRUE` and initiate a parallel plan via {future}.
Since this is only an example, parallelization is not enabled here.
Here we will use a simple regression tree as an example learner.
In practice you might want to use a different learner - you can find an overview of available learners [here](https://mlr3extralearners.mlr-org.com/articles/learners/list_learners.html).

```{r}
learner = lrn("regr.rpart")
set.seed(42)
row_ids = sample(1:task$nrow, 500)
learner$train(task, row_ids = row_ids)

print(learner)
```

### Prediction

For prediction `predict_spatial()` is used.
It will return a raster file which contains the predictions.
Users can select which R spatial format the returned raster should have.

```{r, warning=FALSE}
ras = predict_spatial(task, learner, format = "stars")
names(ras) = "cadmium"

print(ras)
```

### Visualization

Finally we can plot the predictions.
The color vector is extract from the viridis color palette via `dput(viridis::viridis_pal()(5))`.

```{r}
plot(ras, col = c("#440154FF", "#443A83FF", "#31688EFF", "#21908CFF", "#35B779FF", "#8FD744FF", "#FDE725FF"))
```
